Mechanized Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2023-11-28 paper-new/intro.tex
changes to layout and prose of paper

% gradual typing, graduality
\subsection{Gradual Typing and Graduality}
One of the principal ways of categorizing type systems of programming languages is
whether they are \emph{static} or \emph{dynamic}.
In static typing, the code is type-checked at compile time, while in dynamic typing,
the type checking is deferred to run-time. Both approaches have benefits: with static typing,
the programmer is assured that if the program passes the type-checker, their program
is free of type errors, and moreover, the equational theory implies that program optimizations are valid.
Meanwhile, dynamic typing allows the programmer to rapidly prototype
their application code without needing to commit to fixed type signatures for their functions.

\emph{Gradually-typed languages} \cite{siek-taha06, tobin-hochstadt06} allow
for both static and dynamic typing disciplines to be used in the same codebase,
and support smooth interoperability between statically-typed and dynamically-typed code.
This flexibility allows programmers to begin their projects in a dynamic style and
enjoy the benefits of dynamic typing related to rapid prototyping and easy modification
while the codebase ``solidifies''. Over time, as parts of the code become more mature
and the programmer is more certain of what the types should be, the code can be
\emph{gradually} migrated to a statically typed style without needing to
rewrite the project in a completely different language.

%Gradually-typed languages should satisfy two intuitive properties.
The following two properties have been identified as useful for gradually typed languages.
First, the interaction between the static and dynamic components of the codebase
should preserve the guarantees made by the static types.
In particular, while statically-typed code can error at runtime in a gradually-typed language,
such an error can always be traced back to a dynamically-typed term that
violated the typing contract imposed by statically typed code.
% In other words, in the statically-typed portions of the codebase, type soundness must be preserved.
Second, gradually-typed languages should support the smooth migration from dynamic typing
to static typing, in that the programmer can initially leave off the
typing annotations and provide them later without altering the meaning of the
program.

% Graduality property
Formally speaking, gradually typed languages should satisfy the 
\emph{dynamic gradual guarantee}, originally defined by Siek, Vitousek, Cimini,
and Boyland \cite{siek_et_al:LIPIcs:2015:5031}.
This property is also referred to as \emph{graduality} \cite{new-ahmed2018}, by analogy with parametricity.
Intuitively, graduality says that going from a dynamic to static style should not
introduce changes in the meaning of the program.
More specifically, making the types more precise by adding typing annotations
% (replacing $\dyn$ with a more specific type
%such as integers)
will either result in the same behavior as the original, less precise program,
or will result in a type error.

\subsection{Current Approaches to Proving Graduality}
Current approaches to proving graduality include the methods of Abstracting Gradual Typing
\cite{garcia-clark-tanter2016} and the formal tools of the Gradualizer \cite{cimini-siek2016}.
These allow the language developer to start with a statically typed language and derive a
gradually typed language that satisfies the gradual guarantee. The main downside to
these approaches lies in their inflexibility: since the process in entirely mechanical,
the language designer must adhere to the predefined framework.
Many gradually typed languages do not fit into either framework, e.g., Typed Racket
\cite{tobin-hochstadt06, tobin-hochstadt08}.
%
Furthermore, while these frameworks do prove
graduality of the resulting languages, they do not show the correctness of the equational theory,
which is equally important to sound gradual typing. For example, programmers often refactor their
code, and in so doing they rely implicitly on the validity of the laws in the equational theory.
Similarly, correctness of compiler optimizations rests on the validity of the corresponding equations
from the equational theory. It is therefore important that the languages that claim to be gradually typed
have provably correct equational theories.

@@

% Proving graduality via LR
New and Ahmed \cite{new-ahmed2018}
have developed a semantic approach to specifying type dynamism in terms of
\emph{embedding-projection pairs}, which allows for a particularly elegant formulation of the
gradual guarantee.
Moreover, their axiomatic account of program equivalence allows for type-based reasoning
about program equivalences.
%
In this approach, a logical relation is constructed and used to show that the equational theory
is sound with respect to the operational semantics.
%and shown to be sound with respect to
%the notion of observational approximation that specifies when one program is more precise than another.
The downside of this approach is that each new language requires a different logical relation
to prove graduality. Furthermore, the logical relations tend to be quite complicated due
to a technical requirement known as \emph{step-indexing}, where the stepping behavior of terms
must be accounted for in the logical relation.
As a result, developments using this approach tend to require vast effort, with the
corresponding technical reports having 50+ pages of proofs.
%
Additionally, while the axioms of gradual type theory are compositional at a ``global'' level,
they do not compose in the step-indexed setting. One of the main goals of the present work
is to formulate a composable theory of gradual typing in a setting where the stepping behavior
is tracked.

An alternative approach, which we investigate in this paper, is provided by
\emph{synthetic guarded domain theory}.
The techniques of synthetic guarded domain theory allow us to internalize the
step-index reasoning normally required in logical relations proofs of graduality,
ultimately allowing us to specify the logical relation in a manner that looks nearly
identical to a typical, non-step-indexed logical relation.

@@

\subsection{Overview of Remainder of Paper}

In Section \ref{sec:overview}, we give an overview of the gradually-typed lambda
calculus and the graduality theorem.
%
In Section \ref{sec:technical-background}, we provide technical background on gradually typed languages and
on synthetic guarded domain theory.
% 
In Section \ref{sec:gtlc-terms}, we introduce the gradually-typed cast calculus
for which we will prove graduality. We give a semantics to the terms
using the tools of guarded type theory.
% Important here are the notions of syntactic
% type precision and term precision. For reasons we describe below, we
% introduce two related calculi, one in which the stepping behavior of terms is
% implicit (an \emph{extensional} calculus, $\extlc$), and another where this behavior
% is made explicit (an \emph{intensional} calculus, $\intlc$).
%

In Section \ref{sec:gtlc-precision}, we define the term precision ordering
and describe our approach to assigning a denotational semantics to this ordering.
This approach builds on the semantics constructed in the previous section,
but extending it to the term ordering presents new challenges.

% In Section \ref{sec:domain-theory}, we define several fundamental constructions
% internal to SGDT that will be needed when we give a denotational semantics to
% the gradual lambda calculus.
%This includes the Lift monad, predomains and error domains.
%
% In Section \ref{sec:semantics}, we define the denotational semantics for the
% intensional gradually-typed lambda calculus using the domain theoretic
% constructions defined in the previous section.
%
In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
extensional gradual lambda calculus.
%
In Section \ref{sec:discussion}, we discuss the benefits and drawbacks to our approach in comparison
to the traditional step-indexing approach, as well as possibilities for future work.

\section{Overview}\label{sec:overview}

@@

In this paper, we will utilize SGDT techniques to prove graduality for a particularly
simple gradually-typed cast calculus, the gradually-typed lambda calculus.
This is the usual simply-typed lambda calculus with a dynamic type $\dyn$ such that
$A \ltdyn\, \dyn$ for all types $A$, as well as upcasts and downcasts between any types
$A$ and $B$ such that $A \ltdyn B$. The complete definition will be provided in
Section \ref{sec:GTLC}.
The graduality theorem is shown below.

@@

\begin{theorem}[Graduality]
  If $\cdot \vdash M \ltdyn N : \nat$, then
  \begin{enumerate}
    \item $M \Downarrow$ iff $N \Downarrow$.
    \item If $M \Downarrow v_?$ and $N \Downarrow v'_?$ then either $v_? = \mho$, or $v_? = v'_?$.
  \end{enumerate}
\end{theorem}

Details can be found in later sections, but we provide a brief explanation of the terminology and notation:

\begin{itemize}
  \item $M \ltdyn N : \nat$ means $M$ and $N$ are terms of type $\nat$ such that
  $M$ is ``syntactically more precise'' than $N$, or equivalently, $N$ is 
  ``more dynamic'' than $M$. Intuitively this means that $M$ and $N$ are the
  same except that in some places where $M$ has explicit typing annotations,
  $N$ has $\dyn$ instead. The relation $\ltdyn$ on terms is called ``term precision''.
  Term precision is intended to model the situation where a term $M$ behaves the
  same as another term $N$ except that $M$ may error more than $N$.
  Term precision is defined by a set of axioms that capture this intuitive notion.
  This will be described in more detail in Section \ref{sec:GTLC}.

  \item $\cdot \Downarrow$ is a relation on terms that is defined such that $M \Downarrow$ means
  that $M$ terminates, either with a run-time error or a value $n$ of type $\nat$.

  \item $\mho$ is a syntactic representation of a run-time type error, which
  happens, for example, when a programmer tries to call a function with a value whose type
  is found to be incompatible with the argument type of the function.

>>>>

% gradual typing, graduality
\subsection{Gradual Typing and Graduality}
In modern programming language design, there is a tension between \emph{static} typing
and \emph{dynamic} typing disciplines.
With static typing, the code is type-checked at compile time, while in dynamic typing,
the type checking is deferred to run-time. Both approaches have benefits: with static 
typing, the programmer is assured that if the program passes the type-checker, their
program is free of type errors, and moreover, soundness of the equational theory implies
that program optimizations are valid.
Meanwhile, dynamic typing allows the programmer to rapidly prototype
their application code without needing to commit to fixed type signatures for their
functions. Unfortunately, most languages choose between static or dynamic typing.
As a result, programmers that initially write their code in a dynamically typed language
in order to benefit from faster prototyping and development time will need to rewrite the
entire codebase in a static language if they would like to receive the benefits of static
typing once their codebase has matured.

\emph{Gradually-typed languages} \cite{siek-taha06, tobin-hochstadt06} seek to resolve this tension
by allowing for both static and dynamic typing disciplines to be used in the same codebase,
and by supporting smooth interoperability between statically-typed and dynamically-typed code.
This flexibility allows programmers to begin their projects in a dynamic style and
enjoy the benefits of dynamic typing related to rapid prototyping and easy modification
while the codebase ``solidifies''. Over time, as parts of the code become more mature
and the programmer is more certain of what the types should be, the code can be
\emph{gradually} migrated to a statically typed style without needing to
rewrite the project in a completely different language.

%Gradually-typed languages should satisfy two intuitive properties.
% The following two properties have been identified as useful for gradually typed languages.

In order for this to work as expected, gradually-typed languages should allow for
different parts of the codebase to be in different places along the spectrum from
dynamic to static, and allow for those different parts to interact with one another.
Moreover, gradually-typed languages should support the smooth migration from
dynamic typing to static typing, in that the programmer can initially leave off the
typing annotations and provide them later without altering the meaning of the program.

% Sound gradual typing
Furthermore, the parts of the program that are written in a dynamic style should soundly
interoperate with the parts that are written in a static style.
That is, the interaction between the static and dynamic components of the codebase
should preserve the guarantees made by the static types.
In particular, while statically-typed code can error at runtime in a gradually-typed language,
such an error can always be traced back to a dynamically-typed term that
violated the typing contract imposed by statically typed code.

% Moreover, gradually-typed languages should allow for
% different parts of the codebase to be in different places along the spectrum from
% dynamic to static, and allow for those different parts to interact with one another.
% In a \emph{sound} gradually-typed language,
% this interaction should respect the guarantees made by the static types.

% Graduality property
Formally speaking, gradually typed languages should satisfy the 
\emph{dynamic gradual guarantee}, originally defined by Siek, Vitousek, Cimini,
and Boyland \cite{siek_et_al:LIPIcs:2015:5031}.
This property is also referred to as \emph{graduality} \cite{new-ahmed2018},
by analogy with parametricity.
Intuitively, graduality says that going from a dynamic to static style should not
introduce changes in the meaning of the program.
More specifically, making the types more precise by adding typing annotations
will either result in the same behavior as the original, less precise program,
or will result in a type error.

\subsection{Current Approaches to Proving Graduality}

\subsubsection{From Static to Gradual}

Current approaches to constructing languages that satisfy the graduality property
include the methods of Abstracting Gradual Typing
\cite{garcia-clark-tanter2016} and the formal tools of the Gradualizer \cite{cimini-siek2016}.
These allow the language developer to start with a statically typed language and derive a
gradually typed language that satisfies the gradual guarantee. The main downside to
these approaches lies in their inflexibility: since the process in entirely mechanical,
the language designer must adhere to the predefined framework.
Many gradually typed languages do not fit into either framework, e.g., Typed Racket
\cite{tobin-hochstadt06, tobin-hochstadt08}.
%
Furthermore, while these frameworks do prove graduality of the resulting languages,
they do not show the correctness of the equational theory, which is equally important
to sound gradual typing.
For example, programmers often refactor their code, and in so doing they rely
implicitly on the validity of the laws in the equational theory.
Similarly, correctness of compiler optimizations rests on the validity of the
corresponding equations from the equational theory. It is therefore important
that the languages that claim to be gradually typed have provably correct
equational theories.

@@

% [Eric] I moved the next two paragraphs from the technical background section
% to here in the intro.
\subsubsection{Classical Domain Models}

New and Licata \cite{new-licata18} developed an axiomatic account of the
graduality relation on cast calculus terms and gave a denotational
model of this calculus using classical domain theory based on
$\omega$-CPOs. This semantics has scaled up to an analysis of a
dependently typed gradual calculus in \cite{asdf}. This meets our
criterion of being a reusable mathematical theory, as general semantic
theorems about gradual domains can be developed independent of any
particular syntax and then reused in many different denotational
models. However, it is widely believed that such classical domain
theoretic techniques cannot be extended to model higher-order store, a
standard feature of realistic gradually typed languages such as Typed
Racket. Thus if we want a reusable mathematical theory of gradual
typing that can scale to realistic programming languages, we need to
look elsewhere to so-called ``step-indexed'' techniques.

\subsubsection{Operational Reduction Proofs}

A series of works \cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
developed step-indexed logical relations models of gradually typed
languages based on operational semantics. Unlike classical domain
theory, such step-indexed techniques are capable of modeling
higher-order store and runtime-extensible dynamic types
\cite{amalsthesis,nonpmetricparamorsomething,new-jamner-ahmed19}. However,
their proof developments are highly repetitive and technical, with
each development formulating a logical relation from first-principles
and proving many of the same tedious lemmas without reusable
mathematical abstractions. Our goal in the current work is to extract
these reusable mathematical principles from these tedious models to
make formalization of realistic gradual languages tractible.

% Alternative phrasing:
\begin{comment}
\subsubsection{Embedding-Projection Pairs}

The series of works by New, Licata, and Ahmed \cite{new-licata18,new-ahmed2018,new-licata-ahmed2019}
develop an axiomatic account of gradual typing involving \emph{embedding-projection pairs}.
This allows for a particularly elegant formulation of the gradual guarantee.
Moreover, their axiomatic account of program equivalence allows for type-based reasoning about program equivalences.

In \cite{new-licata18}, New and Licata construct a denotational model of the axioms of
gradual typing using tools from classical domain theory. The benefit to this approach is that it is a reusable mathematical theory:
general semantic theorems about gradual domains can be developed independent of any particular syntax and then reused in many different denotational models.
Unfortunately, however, it is unclear how to extend these domain-theoretic techniques to incorporate more advanced language features such as higher-order store, a standard feature of realistic gradually typed languages such as Typed
Racket.
Thus, if we want a reusable mathematical theory of gradual typing that can scale to realistic programming languages, we need to look to different techniques.

In \cite{new-ahmed2018,new-licata-ahmed2019}, a more traditional operational semantics for gradual typing is derived from the axioms.
% The axioms are then proven to be sound with respect to this operational semantics by constructing a logical relation.
A logical relations model is constructed and used to prove that the axioms are sound with respect to the operational semantics.
The downside of this approach is that each new language requires a different logical relation
to prove graduality, even though the logical relations for different languages end up sharing many commonalities.
Furthermore, the logical relations tend to be quite complicated due to a technical requirement known as \emph{step-indexing},
where the stepping behavior of terms must be accounted for in the logical relation.
As a result, developments using this approach tend to require vast effort, with the
corresponding technical reports having 50+ pages of proofs.

% In this approach, a logical relation is constructed and used to show that the equational theory
% is sound with respect to the operational semantics.

% Additionally, while the axioms of gradual type theory are compositional at a ``global'' level,
% they do not compose in the step-indexed setting. One of the main goals of the present work
% is to formulate a composable theory of gradual typing in a setting where the stepping behavior
% is tracked.
\end{comment}

An alternative approach, which we investigate in this paper, is provided by
\emph{synthetic guarded domain theory}.
The techniques of synthetic guarded domain theory allow us to internalize the
step-indexed reasoning normally required in logical relations proofs of graduality,
ultimately allowing us to specify the logical relation in a manner that looks nearly
identical to a typical, non-step-indexed logical relation.

@@

\subsection{Outline of Remainder of Paper}

% In Section \ref{sec:overview}, we give an overview of the gradually-typed lambda
% calculus and the graduality theorem.
%
In Section \ref{sec:technical-background}, we provide technical background on synthetic guarded domain theory.
% 
In Section \ref{sec:gtlc-terms}, we formally introduce the gradually-typed cast calculus
for which we will prove graduality. We give a semantics to the terms using
synthetic guarded domain theory.
% Important here are the notions of syntactic
% type precision and term precision. For reasons we describe below, we
% introduce two related calculi, one in which the stepping behavior of terms is
% implicit (an \emph{extensional} calculus, $\extlc$), and another where this behavior
% is made explicit (an \emph{intensional} calculus, $\intlc$).
%

In Section \ref{sec:gtlc-precision}, we define the term precision ordering for
the gradually-typed lambda calculus and describe our approach to assigning a
denotational semantics to this ordering.
This approach builds on the semantics constructed in the previous section,
but extending it to the term ordering presents new challenges.

% In Section \ref{sec:domain-theory}, we define several fundamental constructions
% internal to SGDT that will be needed when we give a denotational semantics to
% the gradual lambda calculus.
%This includes the Lift monad, predomains and error domains.
%
% In Section \ref{sec:semantics}, we define the denotational semantics for the
% intensional gradually-typed lambda calculus using the domain theoretic
% constructions defined in the previous section.
%
In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
gradual lambda calculus.
%
In Section \ref{sec:discussion}, we discuss the benefits and drawbacks to our approach in comparison
to the traditional step-indexing approach, as well as possibilities for future work.

\subsection{Overview of Results}\label{sec:overview}

@@

In this paper, we will utilize SGDT techniques to prove graduality for a particularly
simple gradually-typed cast calculus, the gradually-typed lambda calculus.
Before stating the graduality theorem, we introduce some definitions related to gradual typing.

% Cast calculi
% TODO move this to earlier?
Gradually typed languages are generally elaborated to a \emph{cast calculus}, in which run-time type checking
that is made explicit. Elaboration inserts \emph{type casts} at the boundaries between static and dynamic code.
In particular, cast insertion happens at the elimination forms of the language (e.g., pattern matching, field reference, etc.).
The original gradually typed language that is elaborated to a cast calculus is called the \emph{surface language}.

% Up and down casts
In a cast calculus, there is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that
$A$ is a \emph{more precise} type than $B$.
There a distinguished type $\dyn$, the \emph{dynamic type}, with the property that $A \ltdyn\, \dyn$ for all $A$.
The parts of the code that are dynamically typed will be given type $\dyn$ in the cast calculus.
%
If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
Upcasts always succeed, while downcasts may fail at runtime, resulting in a type error.
Cast calculi have a term $\mho$ representing a run-time type error for any type $A$.

% Syntactic term precision
We also have a notion of \emph{syntactic term precision}.
If $A \ltdyn B$, and $M$ and $N$ are terms of type $A$ and $B$ respectively, we write
$M \ltdyn N : A \ltdyn B$ to mean that
$M$ is ``syntactically more precise'' than $N$, or equivalently, $N$ is 
``more dynamic'' than $M$. Intuitively, this models the situation where $M$ and $N$
behave the same except that $M$ may error more.
Term precision is defined by a set of axioms that capture this intuitive notion.
The specific rules for term precision depend on the specifics of the language, but
many of the rules can be derived from the typing rules in a straightforward way.
% Cast calculi also have a term $\mho$ representing a run-time type error for any type $A$,
% and since
Since $\mho$ always errors, there is a term precision rule stating that $\mho \ltdyn M$ for all $M$.

% GTLC
The gradually-typed lambda calculus is the usual simply-typed lambda calculus with a dynamic
type $\dyn$ such that $A \ltdyn\, \dyn$ for all types $A$, as well as upcasts and downcasts
between any types $A$ and $B$ such that $A \ltdyn B$. The complete definition will be provided in
Section \ref{sec:gtlc-precision}.

With these definitions, we can state the graduality theorem for the gradually-typed lambda calculus.

@@

\begin{theorem}[Graduality]
  Let $\cdot \vdash M \ltdyn N : \nat$. 
  If $M \Downarrow v_?$ and $N \Downarrow v'_?$, then either $v_? = \mho$, or $v_? = v'_?$.
\end{theorem}

Note:
\begin{itemize}

  \item $\cdot \Downarrow$ is a relation between terms that is defined such that $M \Downarrow$ means
  that $M$ terminates, either with a run-time error or a value $n$ of type $\nat$.

### 2023-11-28 paper-new/ordering.tex
changes to layout and prose of paper

\subsection{Term Precision for GTLC}

@@

\begin{mathpar}
  \inferrule*{M \ltdyn_i N : B}
             {\dn c M \ltdyn_i N : c}
\end{mathpar}

If $c$ is inj-arr, then when we downcast $M$ from $dyn$ to $\dyn \ra \dyn$,
semantically this will involve a $\theta$ because the value of type $dyn$
in the semantics will contain a \emph{later} function $\tilde{f}$.
Thus, in order for the right-hand side to be related to the downcast,
we need to insert a delay on the right.
%
The need for delays affects the cast rules involving upcasts as well, because
the upcast for functions involves a downcast on the domain:

\[ \up{c_i \ra c_o}{M} \equiv \lambda (x : B_i). \up{c_o}(M\, (\dn {c_i} x)). \]

@@

\inferrule
  {\delta_\nat : \perte \nat \and \delta_f : \perte (\dyn \ra \dyn)}
  {\pertdyn{\delta_\nat}{\delta_f} : \perte \dyn}

\inferrule
  {\delta_\nat : \pertp \nat \and \delta_f : \pertp (\dyn \ra \dyn)}
  {\pertdyn{\delta_\nat}{\delta_f} : \pertp \dyn}

>>>>

\subsection{Term Precision for GTLC}\label{sec:gtlc-term-precision-axioms}

@@

\begin{mathpar}
  \inferrule*{M \ltdyn_i N : B}
             {\dnc{c}{M} \ltdyn_i N : c}
\end{mathpar}

If $c$ is inj-arr, then when we downcast $M$ from $dyn$ to $\dyntodyn$,
semantically this will involve a $\theta$ because the value of type $dyn$
in the semantics will contain a \emph{later} function $\tilde{f}$.
Thus, in order for the right-hand side to be related to the downcast,
we need to insert a delay on the right.
%
The need for delays affects the cast rules involving upcasts as well, because
the upcast for functions involves a downcast on the domain:

\[ \up{A_i \ra A_o}{B_i \ra B_o}{M} \equiv \lambda (x : B_i). \up{A_o}{B_o}(M\, (\dn {A_i}{B_i} x)). \]

@@

\inferrule
  {\delta_\nat : \perte \nat \and \delta_f : \perte (\dyntodyn)}
  {\pertdyn{\delta_\nat}{\delta_f} : \perte \dyn}

\inferrule
  {\delta_\nat : \pertp \nat \and \delta_f : \pertp (\dyntodyn)}
  {\pertdyn{\delta_\nat}{\delta_f} : \pertp \dyn}

### 2023-11-28 paper-new/technical-background.tex
changes to layout and prose of paper

\subsection{Gradual Typing}

% Cast calculi
In a gradually-typed language, the mixing of static and dynamic code is seamless, in that
the dynamically typed parts are checked at runtime. This type checking occurs at the elimination
forms of the language (e.g., pattern matching, field reference, etc.).
Gradual languages are generally elaborated to a \emph{cast calculus}, in which the dynamic
type checking is made explicit through the insertion of \emph{type casts}.

% Up and down casts
In a cast calculus, there is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that
$A$ is a \emph{more precise} type than $B$.
There a dynamic type $\dyn$ with the property that $A \ltdyn\, \dyn$ for all $A$.
%
If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
Upcasts always succeed, while downcasts may fail at runtime.
%
% Syntactic term precision
We also have a notion of \emph{syntactic term precision}.
If $A \ltdyn B$, and $M$ and $N$ are terms of type $A$ and $B$ respectively, we write
$M \ltdyn N : A \ltdyn B$ to mean that $M$ is more precise than $N$, i.e., $M$ and $N$
behave the same except that $M$ may error more.

@@

\subsection{Operational Reduction Proofs}

\subsection{Classical Domain Models}

New and Licata \cite{new-licata18} developed an axiomatic account of the
graduality relation on cast calculus terms and gave a denotational
model of this calculus using classical domain theory based on
$\omega$-CPOs. This semantics has scaled up to an analysis of a
dependently typed gradual calculus in \cite{asdf}. This meets our
criterion of being a reusable mathematical theory, as general semantic
theorems about gradual domains can be developed independent of any
particular syntax and then reused in many different denotational
models. However, it is widely believed that such classical domain
theoretic techniques cannot be extended to model higher-order store, a
standard feature of realistic gradually typed languages such as Typed
Racket. Thus if we want a reusable mathematical theory of gradual
typing that can scale to realistic programming languages, we need to
look elsewhere to so-called ``step-indexed'' techniques.

A series of works \cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
developed step-indexed logical relations models of gradually typed
languages based on operational semantics. Unlike, classical domain
theory, such step-indexed techniques are capable of modeling
higher-order store and runtime-extensible dynamic types
\cite{amalsthesis,nonpmetricparamorsomething,new-jamner-ahmed19}. However
their proof developments are highly repetitive and technical, with
each development formulating a logical relation from first-principles
and proving many of the same tedious lemmas without reusable
mathematical abstractions. Our goal in the current work is to extract
these reusable mathematical principles from these tedious models to
make formalization of realistic gradual languages tractible.

@@

  Because the dynamic type is modeled as a non-well-founded
  recursive type, their logical relation needs to be paramterized by natural numbers
  to restore well-foundedness. This technique is known as a \emph{step-indexed logical relation}.
  Intuitively, step-indexing makes the steps taken by terms observable.
  %
  Reasoning about step-indexed logical relations
  can be tedious and error-prone, and there are some very subtle aspects that must
  be taken into account in the proofs. Figure \ref{TODO} shows an example of a step-indexed logical
  relation for the gradually-typed lambda calculus.

  In particular, the prior approach of New and Ahmed requires two separate logical
  relations for terms, one in which the steps of the left-hand term are counted,
  and another in which the steps of the right-hand term are counted.
  Then two terms $M$ and $N$ are related in the ``combined'' logical relation if they are
  related in both of the one-sided logical relations. Having two separate logical relations
  complicates the statement of the lemmas used to prove graduality, because any statement that
  involves a term stepping needs to take into account whether we are counting steps on the left
  or the right. Some of the differences can be abstracted over, but difficulties arise for properties %/results
  as fundamental and seemingly straightforward as transitivity.

>>>>

@@

@@

  Because the dynamic type is modeled as a non-well-founded
  recursive type, their logical relation needs to be paramterized by natural numbers
  to restore well-foundedness. This technique is known as a \emph{step-indexed logical relation}.
  Intuitively, step-indexing makes the steps taken by terms observable.
  %
  Reasoning about step-indexed logical relations
  can be tedious and error-prone, and there are some very subtle aspects that must
  be taken into account in the proofs.
  %
  In particular, the prior approach of New and Ahmed requires two separate logical
  relations for terms, one in which the steps of the left-hand term are counted,
  and another in which the steps of the right-hand term are counted.
  Then two terms $M$ and $N$ are related in the ``combined'' logical relation if they are
  related in both of the one-sided logical relations. Having two separate logical relations
  complicates the statement of the lemmas used to prove graduality, because any statement that
  involves a term stepping needs to take into account whether we are counting steps on the left
  or the right. Some of the differences can be abstracted over, but difficulties arise for properties %/results
  as fundamental and seemingly straightforward as transitivity.
